---
title: Open Science tools in our research group
author: Emil Tveden Bjerglund
date: '2017-07-07'
slug: open-science-tools-in-our-research-group
categories: [R, Open Science]
tags: []
bibliography: ../lib.bib
csl: ../csl/american-chemical-society-with-titles-doi-no-et-al.csl
---

```{r setup, include=FALSE}
library(tidyverse)
carnames <- rownames(mtcars)
mtcars <- as_tibble(mtcars) %>% mutate(car = carnames) %>% select(car, mpg, cyl, hp, wt, am)
```


Currently in the [Organic Surface Chemistry group](//surfchem.dk), which I am part of, there are large variations between group members when it comes to the tools used for data analysis. Some people feel most comfortable in spreadsheet programs such as Origin or Excel, while others rely on a mix of Matlab, R, Python or other tools. We use a lot of different experimental techniques in our research, and therefore generate data in a lot of different formats. Different workflows of varying efficiency often makes joint projects more difficult than necessary.

## Open Science
Recently a lot of our research activities has been oriented towards the [SPOMAN Open Science collaboration](//www.spoman-os.org), and the ongoing projects are available on the [Open Science Framework (osf.io)](//osf.io/wudyt). In most cases, the methods and results are thoroughly described for each project. However, since a lot of the data analysis is currently performed in spreadsheet software, the data analysis is not very transparent and reproducible. A recent paper by Lowndes et al. discusses exactly this issue, and describes how the use of [R](//www.r-project.org) and [Github](//github.com) has allowed them to do better science in less time.[@Lowndes2017]

For a while I have been considering how to apply this thinking to our research. I am confident that it could be made more efficient and well-documented if R was adopted as a general tool for data analysis. I will try to give my reasoning for this below.

### Clear and well-documented analysis in R Markdown
If you recieve a spreadsheet from someone, containing some data analysis it can be very hard to decipher the thinking of the original author (and to be honest, analysis done by yourself 6 monts ago, might aswell be done by a complete stranger). Some of the challenges are:

* There is no connection to the original data. How did the data come from your scientific equipment and into the spreadsheet? Was it in any way altered previously?
* What calculations are performed? Which cells contain data and formulas? You have to manually click around to know.
* There is no easy way to reproduce an analysis, without having to enter new formulas, making new assumptions etc.

R provides a very nice feature which can alleviate these challenges, and make an analysis easier to understand and reproduce: [R Markdown documents](//rmarkdown.rstudio.com). _This post is written in R Markdown to demonstrate some of the capabilities (you can go [here](https://github.com/emiltb/emil.tbjerglund.dk/tree/master/content/post) to see the code)_. It makes it easy to alternate between descriptive paragraphs, describing the analysis and intermediate conclusions, and then R code, where scripts are used to load, modify and plot data.

We could for example provide the script for loading a dataset and directly show how a few of the variables are modified. This is demonstrated here for R's built-in `mtcars` dataset.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
dataset <- mtcars %>% 
  mutate(wt = wt*0.45359,      # Convert lbs to tons
         mpg = mpg * 0.42514,  # Convert mpg to km/L
         Transmission = forcats::fct_recode(as.factor(am), 
                                            Automatic = "0", 
                                            Manual = "1")
    )
knitr::kable(head(dataset), caption = "The mtcars dataset contain data on 32 different cars")
```

We can then plot our dataset, in a completely reproducible manner by a few lines of code. Anyone with the datafile and this script, would be able to produce this exact plot.

```{r fig.height=3, message=FALSE, warning=FALSE}
dataset %>%  
  ggplot(aes(x = wt, y = mpg, color = Transmission)) +
  geom_point() +
  labs(x = "Weight (t)", y = "Mileage (km/L)", title = "Fuel economy")
```

### Reuse of work through an R package
Take any given [research project](http://surfchem.dk/category/articles/) in our group, and you will commonly find that we use X-ray Photoelectron Spectroscopy (XPS), Raman spectroscopy, Infrared spectroscopy, Ellipsommetry and various electrochemical techniques all the time. In essence, a handful of techniques produce the large majority of our data. Most of it is initially saved in proprietary formats and treated in the corresponding software, but it can all be exported as plain-text files (.csv, .txt). Wouldn't it be nice to have a common framework to analyse and plot all data, for every project, regardless of the technique used? This is also possible in R by creating a package!

> Packages are the fundamental units of reproducible R code. They include reusable R functions, the documentation that describes how to use them, and sample data. [**R packages** by Hadley Wickham](//r-pkgs.had.co.nz/)

Initially, an R package for our group could contain functions to load the data from the above-mentioned techniques and clean it up, so that it is ready for analysis. We could then extend on this by providing often-used analysis as functions, shortcuts for making commonly used plots etc. In this way code could easily be reused among members of the group, making our science more efficient. Having such a toolbox provided will also make it easier for new group members to get started with R.

### Collaborative todo-lists and dicussions on github


## Barriers for doing this


### References
